\documentclass[amsmath,amsfonts,rmp,letterpaper]{revtex4}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{silence}
% ----------------------------------------------------------------
\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
\hfuzz2pt % Don't report over-full h-boxes if over-edge is small
% NUMBERING ------------------------------------------------------
%\numberwithin{equation}{section}
% MATH MODE COMMANDS ---------------------------------------------
\newcommand{\V}[1]{\mathbf{#1}} % bold vector or matrix
\newcommand{\sub}[1]{_{\text{#1}}} % text subscript
\newcommand{\di}{\partial} % partial derivative
\newcommand{\unit}[1]{\;\mathrm{#1}} % attach units
\newcommand{\kros}{\times} % vector product
\newcommand{\grad}{\boldsymbol{\nabla}} % gradient operator
\newcommand{\divergenz}{\boldsymbol{\nabla}\cdot} % divergence operator
\newcommand{\curl}{\boldsymbol{\nabla}\kros} % curl (rot) operator
\newcommand{\abs}[1]{\left\vert#1\right\vert} % generic absolute value
\newcommand{\set}[1]{\left\{#1\right\}} % put elements between { }
\newcommand{\Real}{\mathbb R} % Real numbers field
\newcommand{\Cmplx}{\mathbb C} % Complex numbers field
\newcommand{\eps}{\varepsilon} % variirtes epsilon
\newcommand{\mean}[1]{\langle #1 \rangle} % physics mean
\newcommand{\about}{\sim\!}
\newcommand{\arr}{\V{r}}
\newcommand{\arp}{\V{r'}}
\newcommand{\om}{\omega}
\renewcommand{\inf}{\infty}
\newcommand{\const}{\mathrm{const.}}
\newcommand{\red}[1]{\textbf{\textcolor{red}{#1}}}
%\DeclareMathOperator{\XXX}{xxx} % \int like operators
%\DeclareMathOperator*{\XXX}{xxx} % \lim like operators
% HYPER REF AND PDFTEX OPTIONS ----------------------------------------------
\usepackage[dvipsnames,usenames]{color}
\usepackage[pdftex,bookmarks=false,pdfstartview=FitH,colorlinks]{hyperref}
\hypersetup{linkcolor=Sepia,citecolor=Sepia}% additional options
\pdfinfoomitdate 1
% LENGTH PARAMETERS ----------------------------------------------
\addtolength{\textwidth}{-0.8in}% the default in one column is too wide
\addtolength{\oddsidemargin}{0.4in}
%\setlength{\parskip}{0.5ex plus 0.2ex minus 0.2ex}
% ----------------------------------------------------------------
\begin{document}

\title{Theory of figures}%
\author{}
\begin{abstract}
Summary of the theory of figures, providing the mathematical basis for using
gravity measurements to learn about planetary interiors. Mathematical statements
are checked to my satisfaction unless otherwise noted (important exception is the
addition theorem). The primary reference is \citep{Zharkov1978}.
\end{abstract}
\maketitle
\tableofcontents

\section{Definitions and notation}\label{sec:definitions}
Given a density distribution $\rho(\arp)$ inside the planet, the \emph{total
potential} is defined as
\begin{equation}
U(\arr) = V(\arr) + Q(\arr)
\end{equation}
where
\begin{equation}\label{eq:gravpot}
V(\arr) = G\int{\rho(\arp)/\abs{\arr - \arp}\,d\tau'}
\end{equation}
is the gravitational potential and
\begin{equation}\label{eq:cetrifugal_potential}
Q(\arr) = \frac{1}{2}\omega^2r^2\sin^2\theta
\end{equation}
is the centrifugal potential; we will use $d\tau$ for a volume element, $\omega$
is the angular rotation velocity (assumed constant and usually about a principal
axis), $r=\abs{\arr}$, and $\theta$ is the angle from the rotation axis.
\emph{Note that Zharkov and Trubitsyn use an unusual positive
potential, presumably for algebraic convenience. This also means accelerations are
given by the positive gradient of potential.}

When the planet is in hydrostatic equilibrium the level surfaces of potential are
also level surfaces of pressure and of density. Including the free surface where
the pressure $p=0$. This can be shown rigorously \citep[e.g.][]{Batchelor1967} and
is also fairly intuitive. Finding the shape of the planet is therefore reduced to
the problem of finding the level surfaces $U(\arr)=\text{constant}$. This problem
is easy to state but hard to solve, essentially because the volume of space over
which the integral above is taken is unknown and must be found as part of the
self-consistent solution.

Equilibrium figures which differ only slightly from spheres are called \emph
{spheroids}. A dimensionless parameter describing the importance of rotation is
\begin{equation}\label{eq:smallq}
q = \frac{\omega^2a^3}{GM}
\end{equation}
where $M$ is the planet's mass and $a$ is the \emph{equatorial radius}.
% section definitions (end)

\section{Decomposition of potential into spherical harmonics}
\label{sec:spherical_harmonics}
The expression for the gravitational potential, eq.~(\ref{eq:gravpot}), can be
written as a sum of powers of $r$ using the decomposition of $1/\abs{\arr-\arp}$
in Legendre polynomials (see \textsc{legendre.pdf}):
\begin{equation}\label{eq:lege_exp}
\begin{split}
\frac{1}{\abs{\arr-\arp}} &= \frac{1}{r\sqrt{1 - 2t(r'/r) + (r'/r)^2}} = \\
&=
\begin{cases}
\frac{1}{r}\sum_{n=0}^{\infty}(\frac{r'}{r})^nP_n(t),&\qquad r>r',\\
\frac{1}{r}\sum_{n=0}^{\infty}(\frac{r'}{r})^{-n-1}P_n(t),&\qquad r<r',
\end{cases}
\end{split}
\end{equation}
where $\gamma$ is the angle between the radius vectors $\arr$ and $\arp$ and
$t=\cos\gamma$. If $r>r'$ for all points where $\rho(\arp)>0$ (i.e.~inside the
planet) then the potential is called \emph{external}. The Legendre polynomials are
given by Rodrigues's formula:
\begin{equation}\label{eq:rodrigues}
P_n(x) = \frac{1}{2^nn!}\frac{d^n}{dx^n}(x^2 - 1)^n
\end{equation}
(although this is not the easiest way to obtain them explicitly) for
$x\in{[-1,1]}$ and luckily there is no confusion about normalization. It will be
useful to remember that
\begin{equation*}
P_0 = 1, \qquad P_2(t) = \frac{3}{2}t^2 - \frac{1}{2},
\qquad\int_{-1}^{1}P_{n>0}(t')\,dt' = 0.
\end{equation*}
and that $P_n(1)=1$.
The gravitational potential in terms of Legendre polynomials is
\begin{equation}\label{eq:lege_gravity}
V(\arr) = \frac{G}{r}\sum_{n=0}^\infty\int\rho(\arp)P_n(t)(r'/r)^\alpha\,d\tau'
\end{equation}
where
\begin{equation*}
\alpha=
\begin{cases}
n,&\qquad r>r',\\
-(n+1),&\qquad r<r',
\end{cases}
\end{equation*}
and the integration is over the (as yet unknown) volume of the planet.

The expansion in Legendre polynomials is compact and neat but it is of little
utility because it does not separate terms arising from the mass distribution from
those arising from the location where the potential is to be evaluated. In
spherical polar coordinates the variable
\begin{equation*}
t = \cos\gamma = \cos\theta\cos\theta' + \sin\theta\sin\theta'\cos(\varphi -
\varphi')
\end{equation*}
mixes the colatitude $\theta$ and longitude $\varphi$ of the integration
variable and the point of measurement. The salvation comes from the \emph{addition
theorem for spherical harmonics} (\red{which I can't derive}):
\begin{equation}\label{eq:addition}
P_n(\cos\gamma) = P_n(\cos\theta)P_n(\cos\theta') +
2\sum_{m=1}^{n}\frac{(n - m)!}{(n + m)!}P_n^m(\cos\theta)P_n^m(\cos\theta')
\cos\bigl[m(\varphi - \varphi')\bigr],
\end{equation}
with the associated Legendre functions
\begin{equation}\label{eq:assoc_lege}
P_n^m(x) = (-1)^m(1 - x^2)^{m/2}\frac{d^m}{dx^m}P_n(x).
\end{equation}
(Unfortunately there is not a universal consensus on the exact form of the $P_n^m$
functions. If a different normalization is used then the following expressions
will all be somewhat different.) With the aid of the addition theorem we can
decompose the potential as\WarningsOff\footnote{Note typo in Z\&T.}\WarningsOn
\begin{equation}\label{eq:spherical_gravity}
\begin{split}
V(\arr) = \frac{G}{r}&\Biggl(
\sum_{n=0}^\infty{}P_n(\cos\theta)\int_\tau{}\rho(\arp)P_n(\cos\theta')
(r'/r)^\alpha\,d\tau' \\
&+ \sum_{n=1}^{\infty}\sum_{m=1}^{n}P_n^m(\cos\theta)\cos(m\varphi)\int_\tau\frac
{2(n - m)!}{(n + m)!}\rho(\arp)P_n^m(\cos\theta')\cos(m\varphi')\Bigl(\frac
{r'}{r}\Bigr)^\alpha\,d\tau' \\
&+ \sum_{n=1}^{\infty}\sum_{m=1}^{n}P_n^m(\cos\theta)\sin(m\varphi)\int_\tau\frac
{2(n - m)!}{(n + m)!}\rho(\arp)P_n^m(\cos\theta')\sin(m\varphi')\Bigl(\frac
{r'}{r}\Bigr)^\alpha\,d\tau'
\Biggr)
\end{split}
\end{equation}
with $\alpha$ as before.

The above expansion is general, not requiring hydrostatic equilibrium or principal
axis rotation. If the planet is fluid then at equilibrium the rotation will always
be about a principal axis. If we take the polar axis (call it the $z$ axis) to
coincide with the rotation axis then symmetry requires that $\rho(\arr)$ and
therefore $V(\arr)$ cannot depend on the longitude $\varphi$ and that
$\rho(r,\theta)$ must include only even powers of $\cos\theta$ (for symmetry about
the equator). In this case we can write a simpler expansion involving only
ordinary Legendre polynomials of only even degree:
\begin{subequations}\label{eq:simp_sphe_gravity}
\begin{equation}
V(r,\theta)=\frac{G}{r}\sum_{n=0}^{\infty}\left(r^{-2n}D_{2n} + r^{2n +
1}D'_{2n}\right)P_{2n}(\cos\theta)
\end{equation}
with
\begin{align}
D_n &= \int_{r'<r}\rho(\arp)(r')^nP_n(\cos\theta')\,d\tau',\\
D'_n &= \int_{r'>r}\rho(\arp)(r')^{-n-1}P_n(\cos\theta')\,d\tau'.
\end{align}
\end{subequations}
The coefficients $D_n$ are usually replaced with the non-dimensional coefficients
$J_n=D_n/(Ma^n)$.

% section spherical_harmonics (end)

\section{The external potential}\label{sec:external_potential}

If the potential is to be evaluated at a point away from the surface of the
planet then $r>r'$ for all differential volume elements in the integral
expressions above. The general form eq.~(\ref{eq:spherical_gravity}) can be
rearranged slightly and rewritten in a form more convenient for comparison with
measured quantities:
\begin{subequations}\label{eq:external_potential}
\begin{equation}
\begin{split}
V_e = \frac{GM}{r}\Bigg(1 &- \sum_{n=1}^{\infty}(a/r)^nJ_nP_n(\cos\theta)\;+ \\
&+\sum_{n=1}^{\infty}\sum_{m=1}^{n}(a/r)^nP_n(\cos\theta)
\Bigl[C_{nm}\cos(m\varphi) + S_{nm}\sin(m\varphi)\Bigr]\Biggr),
\end{split}
\end{equation}
with the coefficients
\begin{align}
Ma^nJ_n &= -\int\rho(\arp)(r')^nP_n(\cos\theta')\,d\tau' = -D_n,\\
Ma^nC_{nm} &= \frac{2(n - m)!}{(n + m)!}\int\rho(\arp)(r')^nP_n^m(\cos\theta')
\cos(m\varphi')\,d\tau',\\
Ma^nS_{nm} &= \frac{2(n - m)!}{(n + m)!}\int\rho(\arp)(r')^nP_n^m(\cos\theta')
\sin(m\varphi')\,d\tau'.
\end{align}
\end{subequations}
Remember, $M=\int\rho\,d\tau'$ is the planet's mass and $a$ is the equatorial
radius.

(Different normalizations of the associated Legendre functions are often used
leading to expansions with the same form but different meaning of the $J_n$,
$C_{nm}$, and $S_{nm}$ coefficients. There is no easy way to guard against errors
or guess which normalization was used if it is not explicitly given. The
decomposition can also be carried out with the complex form of the Legendre
functions leading to even more confusion.)

A natural choice of reference frame can eliminate some of the coefficients in the
expansion~(\ref{eq:external_potential}). First, if the origin of coordinates is
chosen at the center of mass of the planet, $\V{R}=(X,Y,Z)$, then
\begin{align*}
-MaJ_1 &= \int{z'\,dm'} = MZ = 0,\\
MaC_{11} &= \int{x'\,dm'} = MX = 0,\text{ and}\\
MaS_{11} &= \int{y'\,dm'} = MY = 0
\end{align*}
so that $J_1$, $C_{11}$, and $S_{11}$ vanish. Next, we can relate the expansion
coefficients to the moments of inertia. We designate those:
\begin{subequations}
\begin{align}
B &= \int_\tau\rho(\arp)(x'^2 + z'^2)\,d\tau',\\
A &= \int_\tau\rho(\arp)(y'^2 + z'^2)\,d\tau',\\
C &= \int_\tau\rho(\arp)(x'^2 + y'^2)\,d\tau',
\end{align}
for the principal moments and
\begin{align}
D &= \int_\tau\rho(\arp)\,y'z'\,d\tau',\\
E &= \int_\tau\rho(\arp)\,x'z'\,d\tau',\\
F &= \int_\tau\rho(\arp)\,x'y'\,d\tau',
\end{align}
for the diagonal, so-called \emph{products of inertia}, also called
\emph{centrifugal moments}.
\end{subequations}
The relations with the expansion coefficients in eq.~(\ref{eq:external_potential})
become clear by writing out the degree 2 Legendre functions in Cartesian
coordinates. It is easy to show by direct comparison that
\begin{subequations}
\begin{equation}
-a^2MJ_2 = \frac{A + B}{2} - C,\quad\text{and}\quad a^2MC_{22} = \frac{B - A}{4},
\end{equation}
and also that
\begin{equation}
D = a^2MS_{21},\quad{}E = a^2MC_{21},\quad\text{and}\quad{}F = 2a^2MS_{22}.
\end{equation}
\end{subequations}
Usually $A=B<C$ so that $a^2MJ_2=(C-A)$. Also, if we align the coordinate axes
with the planet's principal axes of inertia then the centrifugal moments vanish
and
\begin{equation}
S_{21} = C_{21} = S_{22} = 0.
\end{equation}
And of course it is still true that for a fluid planet at equilibrium the density
must be independent of longitude and symmetric about the equator and therefore
\begin{equation}
D_{2n+1} = J_{2n+1} = 0,\quad\forall{n}.
\end{equation}

Finally, expressing the centrifugal potential~(\ref{eq:cetrifugal_potential}) in
terms of $P_2$:
\begin{equation}
Q(\arr) = \frac{1}{3}\omega^2r^2\bigl[1 - P_2(\cos\theta)\bigr]
\end{equation}
and using the small parameter $q$ (\ref{eq:smallq}) the total external potential
is
\begin{equation}\label{eq:total_external_potential}
V_e(\arr) = \frac{GM}{r}\left[1 - (a/r)^2J_2P_2 - (a/r)^4J_4P_4 -
(a/r)^6J_6P_6 - \cdots + \frac{1}{3}(r/a)^3\bigl(1 - 
P_2(\cos\theta)\bigr)\,q\right].
\end{equation}

% section external_potential (end)

\section{Level surfaces -- preliminary results}\label{sec:level_surfaces_prelims}
For a slowly rotating planet all level surfaces (surfaces of constant potential)
differ only slightly from spheres. We would like to describe them in the form
\begin{equation}\label{eq:lvl_surf}
r(\theta)=l\bigl[1 + l_{2n}(l)P_{2n}(t)\bigr].
\end{equation}
The coefficients $l_{2n}$ are of order $q^n$ (\red{WHY?}) and $l$ is some
characteristic length. Note that here I begin to use an implicit summation
convention\WarningsOff\footnote{The summation is on $n\ge{0}$. The constant, $n=0$
coefficients are not of order one though because of the explicit $+1$ in
eq.~\eqref{eq:lvl_surf}. An odd choice, I think, but whatever.}\WarningsOn. Also
remember that $t=\cos\theta$.

We can get some preliminary relations simply by plugging in either the equatorial
radius ($a$), the polar radius ($b$) or the mean radius ($s$), into
eq.~(\ref{eq:lvl_surf}), with $\theta=0$ or $\theta=\pi/2$. When we use the
equatorial radius $a$ we denote the coefficients $l_n=a_n$ and similarly for the
other variables. For example, since $a=r(\pi/2)$ we know that
\begin{equation}
a_{2n}(a)P_{2n}(0)=0
\end{equation}
(implicit summation!) and, since $b=r(0)$,
\begin{equation}
\sum_{n=0}^{\infty}b_{2n}(b)=0.
\end{equation}
Now, with $l=a$ and $\theta=0$ we find that
\begin{equation}
b = a\bigl[1 + \sum_{n=0}^{\infty}a_{2n}\bigr]
\end{equation}
and we can describe the \emph{flattening} of a level surface:
\begin{equation}
e(a) = \frac{a - b}{a} = -\sum_{n=0}^{\infty}a_{2n}.
\end{equation}

We will also often use the mean radius of a level surface, $s$, defined by
\begin{equation}
4\pi{}s^3/3 = \int\int\int{}r^2(\theta)\,dr\,d\theta\,d\phi =
(2\pi/3)\int_{-1}^{1}r^3(\theta)\,dt.
\end{equation}
Again, playing the substitution games, we can find several relations:
\begin{subequations}\label{eq:radii_relations}
\begin{equation}
a/s = 1 + s_{2n}P_{2n}(0) = 1 + s_0 - \frac{1}{2}s_2 + \frac{3}{8}s_4 -
\frac{5}{16}s_6 + \cdots,
\end{equation}
\begin{equation}
b/s = 1 + \sum_{n=0}^{\infty}s_{2n},
\end{equation}
and
\begin{equation}
e(s) = \frac{\sum{}s_{2n}[P_{2n}(0) - 1]}{1 + \sum{}s_{2n}P_{2n}(0)}.
\end{equation}
\end{subequations}
The coefficients $s_n$ themselves are coupled by substituting
eq.~(\ref{eq:lvl_surf}) in the definition of $s$, to get
\begin{equation}
\int_{-1}^{1}\bigl[1 + s_{2n}P_{2n}(t)\bigr]^3\,dt = 2.
\end{equation}

Finally, we will sometimes use a small parameter $m=\omega^2{}s^3/GM$ instead of
the parameter $q=\omega^2{}a^3/GM$. The advantage of $m$ is that
\begin{equation}
m = \frac{3\omega^2}{4\pi{G}\bar{\rho}}
\end{equation}
where $\bar{\rho}$ is the average density of the planet. We find the relation
between $m$ and $q$ by using eqs.~(\ref{eq:radii_relations}):
\begin{equation}
q/m = (a_1/s_1)^3 = \bigl[1 + s_{2n}(s)P_{2n}(0)\bigr]^3 = 1 + 3s_0 - \frac{3}
{2}s_2 + \frac{3}{4}s_2^2 + \frac{9}{8}s_4 + \cdots.
\end{equation}
The index $1$ on the coefficients refers to the outermost level surface, i.e., the
surface of the planet.



% section level_surfaces_preliminary_results (end)

% References
\bibliographystyle{plainnat} % [plainnat|abbrvnat|agu08|agufull08|...]
\bibliography{/users/naor/documents/library}

% ----------------------------------------------------------------
\end{document}
% ----------------------------------------------------------------
%END OF FILE
%-----------------------------------------------------------------
